---
title: "Class Reference"
author: "Student name"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.


## Codes to remember 
bind_rows(list(- combines multiple datasets. 
cleaned_conowingo %>% 
  get_dupes(): duplicate rows

command shift m for and then

cleaned_conowingo <- conowingo %>%
  clean_names(): standardizes columns
  cleaned_conowingo <- conowingo %>%
  clean_names() %>% 
  rename(linenumber = x1_linenumber) (adds x in front of column thats starts with an number)
  
True:keeps everything else the same

as.numeric:

Head: shows first few rows

Summary: adds new column, use after group by, used for counting, used for caculations

mutate:adds new column to clean data, does not do any math

guess_max: guess data types on first 10 rows

Colnames:name of columns

Glimpse: to get a sense of the data, some of the values in each field/variable/column, and the data types of each

# cleaning function
cleaned_conowingo <- conowingo %>%
  clean_names() %>% 
  rename(linenumber = x1_linenumber) %>% 
  mutate(amount = as.numeric(amount)) %>% 
  distinct(): gets rid of duplicate rows
# display the cleaned dataset
cleaned_conowingo

cleaned_conowingo <- conowingo %>%
  clean_names() %>% 
  rename(linenumber = x1_linenumber) %>% 
  mutate(amount = as.numeric(amount)) %>% 
  distinct() %>%
  mutate(zip = str_sub(zip, start=1L, end=5L)) :standardizes zip code
  

  mutate(city = str_to_title(city)) :first letter uppercase
  
   mutate(city = case_when(
    str_detect(city,"^Conowing") ~ "Conowingo",
    TRUE ~ city : fixes misspellings
  
 To get percent: select(office, district, name_raw, party, jurisdiction, election_day, votes) %>% 
  mutate(
  percent_election_day = (election_day/votes)*100
  case_when: Each line is basically a filter followed by `~` and then a value for the new column for records that match that filter
  maryland_winred_categories <- maryland_winred %>%
  mutate(
    amount_category = case_when(
        amount < 100 ~ "Less than $100",
        amount >= 100 & amount <= 500 ~ "Between $100 and $500",
        amount > 500 & amount <= 2900 ~ "Between $501 and $2900",
        amount > 2900 ~ "More than $2,900"
      )
  )
head(maryland_winred_categories)

```{r}
setwd("~/Documents/GitHub/data_journalism_2022_spring")
```

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### Intersting Packages
Lubridate: Make Dealing with Dates a Little Easier. Functions to work with date-times and time-spans: fast and user friendly parsing of date-time data, extraction and updating of components of a date-time (years, months, days, hours, minutes, and seconds), algebraic manipulation on date-time and time-span objects.
### Summarizing

I need to use group_by and summariz. Here's an example of grouping by county and calculating counts, sum and other descriptive statistics.


```{r}
ppp_maryland_loans %>%
  group_by(project_county_name) %>%
  summarise(
    count_loans = n(),
    total_loans_amount = sum(amount),
    mean_loan_amount = mean(amount),
    median_loan_amount = median(amount),
    min_loan_amount = min(amount),
    max_loan_amount = max(amount)
  ) %>%
  arrange(desc(max_loan_amount))
```

